%------------------------------------------------------
% Study 1
%------------------------------------------------------
\input{tex/method_files/study1}

%------------------------------------------------------
% Study 2
%------------------------------------------------------
\section{Study 2: Towards a Conceptual Framework for Bridging Player and Learner Roles}

Gamification is highly context-dependent, and ill-designed gamification solutions can lead to harmful effects instead of the expected benefits [4, 5]. Thus, the construction of sound gamified CL sessions requires not only careful analysis of appropriate gamified activities (e.g., environment’s design) and meaningful rewards (e.g., suitable game elements) but also, the ability of assigning appropriate player roles to each learner. Towards filling such a gap, we set out to create a conceptual framework to handle models and vocabulary to represent learner-player roles interactions. 

In Chapter \ref{chapter:background-cscl} we briefly introduced some CSCL concepts , such as group formation and in

and in Chapter \ref{chapter:persuasive-techonology}, we presented  

This chapter will use these concepts and will show the models we devised as well as the vocabulary gathered and organized to represent learner-player roles interactions in gamified CL contexts.

Our approach for creating player roles is based on the Motivations to Play  proposed by [19]. In [19], differently from most available player models, instead of using psychological archetypes in an effort to fit a player in one kind of dominant personality type, the proposed approach tries to identify not only the reasons that can motivate an individual to play a video game,  but also their relationship and overlaps. In this way, the model proposed by [19] does not attempt to identify in which archetype one individual fits, but rather to understand what can motivate such an individual to play. In addition, by identifying the reasons that may arouse motivation to play, by analyzing the results of the scoring system developed, one can also identify what can be deemed less attractive for such an individual. 
To help us better understand player’s needs that drive motivations to play, we rely on Self-determination Theory (SDT) concepts [13]. SDT seeks to explain how intrinsic and extrinsic motivators influence human behavior and the development of individuals. The three basic psychological needs, considered fundamental to influence motivation are autonomy, relatedness, and competence. According to [13], by promoting the internalization of these feelings, individuals have the potential to carry out their activities with improved performance, persistence, and creativity, for instance. 

\begin{table}[]
\centering
\caption{Motivations to play (Yee, 2008) X STD (Ryan and Deci, 2001) X Player role (also Ferro et al., 2013)
}
\label{my-label}
\begin{tabular}{|l|l|c|c|c|l|}
\hline
\multicolumn{2}{|c|}{\textbf{Motivations to play}} & \multicolumn{3}{c|}{\textbf{Self-Determination Theory}} & \multicolumn{1}{c|}{\textbf{Player role}} \\ \hline
Component & Subcomponent & \multicolumn{1}{l|}{Competence} & \multicolumn{1}{l|}{Relatedness} & \multicolumn{1}{l|}{Autonomy} &  \\ \hline
\multirow{3}{*}{Achievement} & Advancement & x &  &  & \multirow{2}{*}{Achiever} \\ \cline{2-5}
 & Mechanics & x &  &  &  \\ \cline{2-6} 
 & Competition & x &  &  & Conqueror \\ \hline
\multirow{3}{*}{Social} & Socializing &  & x &  & \multirow{3}{*}{Humanist} \\ \cline{2-5}
 & Relationship &  & x &  &  \\ \cline{2-5}
 & Teamwork &  & x &  &  \\ \hline
\multirow{4}{*}{Immersion} & Discovery &  &  & x & Explorer \\ \cline{2-6} 
 & Customization &  &  & x & Creator \\ \cline{2-6} 
 & \textit{Role-playing}* &  &  &  &  \\ \cline{2-6} 
 & \textit{Escapism}* &  &  &  &  \\ \hline
\end{tabular}
\end{table}


\begin{table}[!htb]
\centering
\caption{Player roles description}
\label{my-label}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Player role} & \textbf{Description} & \textbf{Source} \\ \hline
\begin{tabular}[c]{@{}l@{}}Achiever\\   (goal-oriented)\end{tabular} & \begin{tabular}[c]{@{}l@{}}They enjoy not only completing a game,
but also being the best winner, accumulating\\
all rewards the game can offer. They are motivated by\\
receiving glory (points, titles, medals, trophies,\\
achievements); gathering (virtual currency and\\
goods); and collecting rare or all in-game items\\
(equipment, weapons, armor, vehicles, mounts, pets).\end{tabular} & {[}19, 24, 26–28{]} \\ \hline
\begin{tabular}[c]{@{}l@{}}Conqueror\\   (people-oriented)\end{tabular} & \begin{tabular}[c]{@{}l@{}}They enjoy rushing and competing against other people.\\ 
Usually, they enjoy testing their skills and seeing\\
how they stack up against other people. They find\\
external ranking systems and zero-sum game mechanics\\
appealing.\end{tabular} & {[}19, 24, 27, 28{]} \\ \hline
\begin{tabular}[c]{@{}l@{}}Humanist\\   (people-oriented)\end{tabular} & \begin{tabular}[c]{@{}l@{}}They enjoy socializing with people (e.g.,\\
being part of groups and teams, forming\\ 
partnerships, and playing collaborative games).\\
They value socializing, sharing learning, and\\
relationship building via shared tasks.\end{tabular} & {[}19, 24, 27, 28{]} \\ \hline
\begin{tabular}[c]{@{}l@{}}Explorer\\   (system-oriented)\end{tabular} & \begin{tabular}[c]{@{}l@{}}They enjoy exploring the system by\\
discovering the ins-and-outs (e.g., hidden or\\
remote places, finding loopholes, knowing the\\
rules that govern a space).\end{tabular} & {[}19, 24, 27, 28{]} \\ \hline
\begin{tabular}[c]{@{}l@{}}Creator\\   (system-oriented)\end{tabular} & \begin{tabular}[c]{@{}l@{}}They enjoy customizing the system (e.g.,\\ 
backgrounds, fonts, buildings, characters,\\
armor, weapons, and vehicles).\end{tabular} & {[}19, 24, 26–28{]} \\ \hline
\end{tabular}
\end{table}

%------------------------------------------------------
% Study 3
%------------------------------------------------------
\section{Study 3: Brazilian Portuguese Cross-Cultural Adaptation and Validation of the Sensibility to Persuasion Scale (Br-STPS)}

%------------------------------------------------------
% Study 4
%------------------------------------------------------
\section{Study 4: Persuasion Profiling to Cope with Learners' Willingness to Join Groups in CSCL}

%-------------------------------------------------------
% Concluding Remarks
%-------------------------------------------------------
\section{Concluding Remarks}