% ---
% Agradecimentos
% ---
I am thankful to Dr. Seiji Isotani for his invaluable feedback and guidance throughout the past years. 
It is utterly my fault if his high standards are not reflected in this document.

My sincere thanks also goes to Dr. Riichiro Mizoguchi. 
During my period abroad I learned a lot from him. 
Our chats helped me to figure out how I should go about doing research and writing papers --- 
In addition, he helped a lot teaching me the "Japanese way" to see the world and to do things (as much as a "gaijin" can learn about the marvelous of this country). And for this I am really in debt. Thank you sensei.
% Ler menos, pensar e escrever mais. Não esquecer de viver, work hard live harder.

\todo[inline] {I am thankful to all members of Ikeda's Lab 

Isotani's Lab. In particular, 

Ikue, Pasimee,  
Grateful thanks are due to Raja. I am glad we became friends. I enjoyed spending most ... with you. Also, it goes without saying that all help during my period in Japan.

For the sake of finishing my Ph.D., I took many friends for granted. Still, they have not given up on me. 
Besides friends, some colleagues and fellow researchers that I met along the way have contributed to make the last few years even more special. The list of friends, colleagues, and fellow researchers is the following (I hope I did not leave anyone off the list)  

I am immensely grateful to my sister~(Adriana), my brother-in-law~(Sebasti\~ao) and nephew (Bruno).
I know that they will never want to read this document but they contributed to it being written. Thank you both for your unconditional support and love.

This research was made possible by the financial support provided by CNPQ (grant number) and CAPES (grant number ).     }